Quantifying the importance of an inference model in Bayesian phylogenetics

### 2020-07-25 pirouette_supplement.tex
Describe pipeline in more detail, as requested by @rsetienne

This subsection shows the diagnostics from the main example, which
uses one tree.
To assess if the results of the inference are meaningful one important 
parameter is the Effective Sample Size (ESS). This quantity describes how 
many independent trees are sampled from the posterior distributions. 
For reliable results it is good practice to have at 
least $ESS = 200$ (see 
\begin{sloppypar}
  \url{https://beast.community/ess_tutorial}).
\end{sloppypar}
In the following we present the ESS for the posterior distributions of 
the 4 cases shown in Fig. \ref{fig:example_30_full_pipeline}: 
"true" pipeline with generative model (Table \ref{tab:esses_true_gen}), 
"true" pipeline with best candidate model (Table \ref{tab:esses_true_best}), 
"twin" pipeline with generative model (Table \ref{tab:esses_twin_gen}), 
"twin" pipeline with best candidate model (Table \ref{tab:esses_twin_best}). 
We also report the marginal likelihood (or evidence) data for model selection 
performed both in the "true" (Table \ref{tab:true_evidence}) and 
"twin" pipeline (Table \ref{tab:twin_evidence}).

>>>>

This subsection describes the pipeline of the main example 
and its diagnostics in more detail.

The pipeline starts at the top-left panel of figure \ref{fig:pipeline} (which 
is identical to figure \ref{fig:dd_tree}),
which is the 'true tree'.
The 'true tree' is generated by the diversity-dependent (DD) 
tree model [\citep{DDD, etienne2012diversity}],
which is a BD model with a speciation rate that is dependent on the number of species,
with (an arbitrarily chosen) crown age of 10 time units 
and 6 tips for an extinction rate of 0.1. 
The carrying-capacity is set to 6. 
The initial speciation rate $\lambda_0$ is chosen 
such that the expected number of species in a constant-rate BD model 
would be equal to the number of tips, which amounts to $\lambda_0 = 0.63$.

From this 'true tree', a 'true alignment' is simulated, using
the JC69 nucleotide substitution model and a strict clock model.
The resulting alignment is shown at the center-left
of figure \ref{fig:pipeline}.

From the 'true alignment' the generative inference model is run.
Of course, it cannot be the actual (DD) model. Instead, the
default BEAST2 inference model is used, which assumes a JC69 nucleotide
substitution model, a strict clock model and a Yule tree model.
The resulting posterior trees are shows in the
center-left panel of figure \ref{fig:pipeline}.

From this 'generative true' posterior (center-left panel in 
figure \ref{fig:pipeline}), the difference between each of its trees is
compared to the 'true tree' (top-left panel), using the nLTT statistic,
resulting in the error distribution shown at the bottom-left panel
of figure \ref{fig:pipeline}.

Based on the 'true alignment' (center-left panel), the candidate model
with the highest marginal likelihood is determined, from a set of 
15 models. The set of models consists out of all combinations
of all 4 nucleotides substitution models (JC69, HKY, TN93, GTR),
all 2 clock models (strict and relaxed log-normal) and 2 birth-death
models (Yule and Birth-Death), except for the inference model used as the
generative model (JC69, strict, Yule). The inference model that had the
highest evidence (as shown in Table \ref{tab:true_evidence}) was 
the inference model with a JC69 nucleotide substitution model,
a relaxed log-normal clock model and a Yule tree model.
From the 'true alignment' this best candidate inference model is run.
The resulting posterior trees are shown in the second panel
of the third row of posteriors in figure \ref{fig:pipeline}.

From this 'best true' posterior (second panel in third row of
figure \ref{fig:pipeline}), the difference between each of its trees is
compared to the 'true tree' (top-left panel), using the nLTT statistic,
resulting in the second error distribution in the bottom row
of figure \ref{fig:pipeline}.

From the 'true tree' (top-left) we generated a BD twin tree (top-right).

From this 'twin tree', a 'twin alignment' is simulated, using
the JC69 nucleotide substitution model and a strict clock model.
The resulting alignment is shown at the center-right
of figure \ref{fig:pipeline}.

From the 'twin alignment' the generative inference model is run as well.
Also here, the default BEAST2 inference model is used, 
which assumes a JC69 nucleotide
substitution model, a strict clock model and a Yule tree model.
The resulting posterior trees are shows in the
third panel of the third row of figure \ref{fig:pipeline}.

From this 'generative twin' posterior (third panel in 
third row of figure \ref{fig:pipeline}), 
the difference between each of its trees is
compared to the 'twin tree' (top-right panel), using the nLTT statistic,
resulting in the error distribution shown at the third panel
of the bottom row of figure \ref{fig:pipeline}.

Based on the 'twin alignment' (center-right panel), the candidate model
with the highest marginal likelihood is determined, from the same 
set of 15 candidate models.The inference model that had the
highest evidence (as shown in Table \ref{tab:twin_evidence}) was 
the inference model with a JC69 nucleotide substitution model,
a strict clock model and a BD tree model (note: this indeed matches how
the twin tree and twin tree have been simulated).
From the 'twin alignment' this best candidate inference model is run.
The resulting posterior trees are shown in the fourth panel
of the third row of posteriors in figure \ref{fig:pipeline}.

From this 'best twin' posterior (fourth in third row of
figure \ref{fig:pipeline}), the difference between each of its trees is
compared to the 'twin tree' (top-right panel), using the nLTT statistic,
resulting in the fourth error distribution in the bottom row
of figure \ref{fig:pipeline}.

To assess if the results of the inference are meaningful one important 
parameter is the Effective Sample Size (ESS). This quantity describes how 
many independent trees are sampled from the posterior distributions. 
For reliable results it is good practice to have at 
least $ESS = 200$ (see 
\begin{sloppypar}
  \url{https://beast.community/ess_tutorial}).
\end{sloppypar}
In the following we present the ESS for the posterior distributions of 
the 4 cases shown in Fig. \ref{fig:example_30_full_pipeline}: 
"true" pipeline with generative model (Table \ref{tab:esses_true_gen}), 
"true" pipeline with best candidate model (Table \ref{tab:esses_true_best}), 
"twin" pipeline with generative model (Table \ref{tab:esses_twin_gen}), 
"twin" pipeline with best candidate model (Table \ref{tab:esses_twin_best}). 
We also report the marginal likelihood (or evidence) data for model selection 
performed both in the "true" (Table \ref{tab:true_evidence}) and 
"twin" pipeline (Table \ref{tab:twin_evidence}).
